---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
We download the data from the source, unzip it and read the activity.csv file: 
```{r loading}
exdata <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(exdata, destfile = "./exdata.zip", method = "curl")
unzip("exdata.zip")
activity <- read.table("activity.csv", header = TRUE, sep = ",")
```

The activity dataframe has the following characteristics:
```{r summary, echo= TRUE}
str(activity)
```

## What is mean total number of steps taken per day?
We first calculate the total number of steps taken per day and store it in a new df
```{r total}
totalsteps<- aggregate(activity$steps, by=list(activity$date), FUN= sum)
names(totalsteps)<- c("Date", "Steps")
```

The new df looks like this:
```{r strtotal, echo= TRUE}
head(totalsteps,3)
```

We now plot the histogram of the total steps per day:
```{r histogram, echo= TRUE, message=FALSE,warning=FALSE}
library(ggplot2)
ggplot(data = totalsteps, aes(Steps)) + geom_histogram(binwidth=1000) +labs(x="Total number of steps", y = "Count of days")+ ggtitle("Histogram of total number of steps")
```

We also calculate the mean and median of the total number of steps taken per day, ignoring NA values:
```{r meanstats, echo= TRUE}
mean_steps <- mean(totalsteps$Steps, na.rm= TRUE)
print(mean_steps)

```

```{r medianstats, echo= TRUE}
median_steps <- median(totalsteps$Steps, na.rm= TRUE)
print(median_steps)

```

## What is the average daily activity pattern?
We first calculate the average number of steps taken per interval and store it in a new dataframe meansteps
```{r totalac}
meansteps<- aggregate(activity$steps, by=list(activity$interval), FUN= mean, na.rm= TRUE)
names(meansteps)<- c("Interval", "Steps")
```

We then make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken (per interval) averaged across all days (y-axis) 
```{r timeseries, echo= TRUE,message=FALSE}
ggplot(data = meansteps, aes(Interval,Steps)) + geom_line() +labs(x="5-minute intervals", y = "Steps")+ ggtitle("Daily activity pattern")
```

The 5-minute interval that, on average, contains more steps is:
```{r max, echo= TRUE}
maxint<- meansteps$Interval[which.max(meansteps$Steps)]
print(maxint)
```

## Imputing missing values
We calculate the count of missing Steps values
```{r missing, echo= TRUE}
NAcount<- sum(is.na(activity$steps))
print(NAcount)
```

Create a new data frame where missing values are replaced with the means for each interval
```{r imputeddf}
library(plyr)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
activity_imputed<- ddply(activity, ~ interval, transform, steps = impute.mean(steps))
```


Calculate the total number of steps taken per day based on imputed data and store it in a new df
```{r totalimputed}
totalimputedsteps<- aggregate(activity_imputed$steps, by=list(activity_imputed$date), FUN= sum)
names(totalimputedsteps)<- c("Date", "Steps")
```

Make a histogram of the imputed data

```{r newhistogram, echo= TRUE}
ggplot(data = totalimputedsteps, aes(Steps)) + geom_histogram(binwidth=1000) +labs(x="Total number of steps", y = "Count of days")+ ggtitle("Histogram of total number of steps, imputed data")
```

We calculate the mean and median of the total number of steps taken per day, with imputed values:
```{r sumstatsmeanimputed, echo= TRUE}
mean_stepsimputed <- mean(totalimputedsteps$Steps, na.rm= TRUE)
print(mean_stepsimputed)
```

```{r sumstatsmedianimputed, echo= TRUE}
median_stepsimputed <- median(totalimputedsteps$Steps, na.rm= TRUE)
print(median_stepsimputed)
```

## Are there differences in activity patterns between weekdays and weekends?
We create a new factor variable in the imputed dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day

```{r wday,message=FALSE, warning=FALSE}
library(lubridate)
activity_imputed$wday <- wday(activity_imputed$date, label = TRUE)
activity_imputed$wday <-  ifelse(activity_imputed$wday == "Sat"| activity_imputed$wday=="Sun", "weekend", "weekday")
```

We first calculate the average number of steps taken per interval grouped by weekend or weekday and store it in a new dataframe meanimputedwdaysteps
```{r avgwday}
meanimputedwdaysteps<- aggregate(activity_imputed$steps, by=list(activity_imputed$interval, activity_imputed$wday), FUN= mean, na.rm= TRUE)
names(meanimputedwdaysteps)<- c("Interval", "Wday", "Steps")
```

Now we make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r panel, echo= TRUE}
ggplot(data = meanimputedwdaysteps, aes(Interval,Steps)) + geom_line() +labs(x="5-minute intervals", y = "Steps")+ ggtitle("Weekday/Weekend activity pattern")+facet_wrap(~Wday,ncol = 1)
```
